---
title: "Crime in Mexico"
output: html_document
date: "2024-10-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Crime Data Analysis in Mexico
This report analyzes crime data in Mexico, providing an overview of the types of crimes and their distribution across time and regions.

### Loading and Preparing Data
The following code loads the dataset and ensures correct encoding for special characters in Spanish.


```{r}
library(readr)
library(tidyverse)
library(dplyr)


# Load data with correct encoding
data_load <- read_csv("CrimeData.csv", locale = locale(encoding = "latin1"), show_col_types = FALSE)
data_load[] <- lapply(data_load, function(x) if (is.character(x)) enc2utf8(x) else x)

head(data_load)
```

### Transforming Data to Long Format
We pivot the data to long format to analyze the monthly crime cases more effectively.


```{r}

# Convert the data to long format
data <- data_load %>%
  pivot_longer(cols = c("Enero", "Febrero", "Marzo", "Abril", "Mayo", "Junio",
                        "Julio", "Agosto", "Septiembre", "Octubre", "Noviembre", "Diciembre"),
               names_to = "Month", values_to = "Cases")

# Drop unnecessary columns and group by crime type
data <- data %>%
  select(-`Bien jurídico afectado`, -`Clave_Ent`, -`Modalidad`) %>%
  group_by(Año, Month, Entidad, `Tipo de delito`) %>%
  summarise(Cases = sum(Cases, na.rm = TRUE), .groups = "drop")

# Display the first few rows to verify
head(data)


```